---
title: "Covid-19 Mortality Analysis"
author: "Felix Reynoso"
date: "5/29/2021"
output: html_document
---

# Introduction
The COVID-19 pandemic, is an ongoing global pandemic of coronavirus disease. The virus was first identified in December 2019 in Wuhan, China. The World Health Organization declared a International Health Emergency on 30 January 2020, and was later declared a pandemic on 11 March 2020. 

In our analysis we explore mortality rates of the COVID-19 virus. We use the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19). With this short description we proceed with our analysis. 

```{r hey, include = FALSE}
# Import needed packages
library(tidyverse)
library(lubridate)
library(zoo)
library(countrycode)
library(moderndive)

global_cases <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")

global_deaths <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")

options(dplyr.summarise.inform = FALSE)
```

# Rate of Spread of COVID-19
The first case of COVID-19 was officially identified on December 2019 in Wuhan, China, and our data goes back to as early as January 2020. 


```{r}
# Adds up the total number of Covid cases each country has had, overtime.
df <- global_cases %>% select(-'Province/State', -'Lat', -'Long', -'Country/Region' ) 
df <- data.frame(date = names(df), total_cases = colSums(df)) %>%
  mutate(date = mdy(date)) %>%
  mutate(year = year(date), month = month(date)) %>% 
  group_by(year, month) %>%
  summarize(cases=max(total_cases)) %>%
  mutate(date=as.yearmon(paste(year, month, sep='-'))) %>%
  ungroup() %>%
  select(date, cases) 

ggplot(df, aes(x = date, y = cases)) + 
  geom_line(color="purple") +
  labs(x = "Time", y = "Number of Confirmed Cases") 
```

```{r, include = FALSE}
f <- df %>% tail(n = 1)

```

From the above graph we can see that the rate of spread of new infections is exponential. As of 
`r f$date`, there have been `r f$cases` confirmed cases of the Coronavirus disease. 


## COVID-19 Mortality Rate

Now we would like to explore the mortality rate of the virus, this rate is the percentage of deaths by cases, which is given by the number deaths divided by the number of confirmed cases.


```{r}
# Plots the Mortality Rates of all countries 
latest <- colnames(global_cases) %>% last()

cases <- global_cases %>% 
  select('Country/Region', all_of(latest)) %>%
  rename(Total_Cases = all_of(latest), Country='Country/Region')

cases <- aggregate(Total_Cases ~ Country, cases, sum)
  
deaths <- global_deaths %>%
  select('Country/Region', all_of(latest)) %>%
  rename(Total_Deaths = all_of(latest), Country = 'Country/Region')

deaths <- aggregate(Total_Deaths ~ Country, deaths, sum)  

df <- inner_join(cases, deaths, by = 'Country') %>%
  mutate(Death_Rate=(Total_Deaths/Total_Cases)*100)

df <- na.omit(df)

global_death_rate <- mean(df$Death_Rate) # Compute the avg global mortality rate

ggplot(df, aes(x = Country, y = Death_Rate)) + 
  geom_col(aes(fill=ifelse(Death_Rate >= global_death_rate,"Above Avg","Below Avg"))) +
  theme(axis.text.x=element_blank()) +
  labs(x="Countries", y = "Mortality Rate", fill="Types") 

```


The global average mortality rate of the COVID-19 is `r global_death_rate`%, but from the above graph we can see that many countries, are way above this average. 

Here are the ten countries with the highest mortality rate.

```{r}
# Plots the 10 counties with highest mortality mate
df1 <- df %>%
  arrange(Death_Rate) %>%
  tail(n=10) 

ggplot(df1, aes(x = Country, y = Death_Rate)) +
  geom_col(fill = "orange") +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("Mortality Rate")
```



```{r, include = FALSE}
# Create a new dataframe, by adding a $code column which is the code of the country, e.g('USA'),
# and uses that column to join the global_cases and global_deaths datasets. Got population data
# from the world_bank_pop database from the tidyverse package. And GDP data from the world bank
# portal, more detail can be found in the '/Data/readme.txt'
df$Code <- countrycode(df$Country, origin='country.name', destination='iso3c')
df <- df %>% select(Country, Death_Rate, Code, Total_Deaths) %>%
  filter(!is.na(Code))

pop <- world_bank_pop %>% 
  filter(indicator == 'SP.URB.TOTL') %>% 
  select(country, colnames(world_bank_pop)[ncol(world_bank_pop)]) %>%
  arrange(country) %>%
  rename(Code = country, Population = '2017')

GDP <- read_csv("Data/countries_GDP.csv") 
GDP <- GDP %>% 
  select('Country Code', colnames(GDP)[ncol(GDP)-2]) %>%
  rename(Code = 'Country Code', GDP = colnames(GDP)[ncol(GDP)-2])
  
df <- df %>% inner_join(pop)
df <- df %>% inner_join(GDP) 
df <- df %>% filter(!is.na(GDP))
df <- df %>% mutate(GDP_Per_Capita = GDP/Population) 
```

## Relationship Between Countries GDP and Mortality Rate
We have seen that mortality rates widely vary by country, this could be due to many factors, but we would like to establish a relationship between the mortality rate of the virus and countries GDP, to examine if there exists any correlation between these two variables. 

```{r}
# Plot a bubble chart, and regression line 
ggplot(df, aes(x = GDP_Per_Capita, y = Death_Rate, size = Population)) + 
  geom_point(alpha=.15, color="black") + 
  scale_size(range = c(.1, 24), name="Population") +
  labs(
    title = "Relationship Between Covid-19 Mortality Rate and Country's GDP",
    x = 'GPD per Capita', 
    y = "Mortality Rate") + 
  geom_smooth(method = "lm", se = FALSE) +
  theme(legend.position = "none")

cor <- df %>% get_correlation(formula = Death_Rate ~ GDP)
```

In the above graph, bubbles are countries and their size represents their population.
We can see that the slope of the regression line is negative, the correlation coefficient between GDP and Mortality Rate is `r cor`%. And this means that countries with higher economic resources, tend to have a lower mortality rate for COVID-19 virus. 


## Bias
Before performing this analysis I had previously watched [The best stats you've ever seen - Hans Rosling](https://www.youtube.com/watch?v=usdJgEwMinM), a Ted talk where the famous statistician Hans Rosling was doing an analysis of the correlation between countries GDPs and child survival rates. I took inspiration from that presentation and thought that if there was a correlation between child survival and GDP, there would probably be a correlation between COVID-19 mortality rates and countries GDP. And it seems a little obvious since generally the higher the GDP, the better the health systems a country has, and therefore the better they can ensure the safety of their citizens. 

This was a possible source of bias I was able to identify during my analysis. 

# Conclusion

In our analysis we've found that:

- Growth of the spread of the COVID-19 virus is exponential
- There are many countries with a mortality rate much higher than `r global_death_rate`%
- There is a negative correlation of `r cor`% between a countries GDP and COVID-19 mortality rate


